---
title: "es1"
output:
  pdf_document: default
  html_document: default
date: "2024-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r message=FALSE, warning=FALSE}
shhh = function(lib_name){ # It's a library, so shhh!
  suppressWarnings(suppressMessages(require(lib_name, character.only = TRUE)))
}
shhh("tidyverse")
shhh("ACutils")
shhh("mvtnorm")
shhh("salso")
shhh("FGM")
shhh("gmp")
shhh("mcclust")
shhh("mcclust.ext")
shhh("logr")
shhh("tidygraph")
shhh("ggraph")
shhh("igraph")
shhh("Rcpp")
shhh("RcppArmadillo")
shhh("RcppEigen")

## Load custom functions
source("functions/utility_functions.R");
source("functions/bulky_functions.R");
source("functions/data_generation.R")
sourceCpp("functions/wade.cpp")
Rcpp::sourceCpp('functions/UpdateParamsGSL.cpp')

library('RcppGSL')
library(fda)
library(tidyverse)
library(coda)
library(lattice)
```

# Goal

Smoothing di dati funzionali simulati, con grafo e partizione tenuti fissi.

# Simulated data

Simuliamo i dati in sim_data.R con p=40 basi e 3 cluster di dimensioni [13, 13, 14]. Grafo e matrice di precisione sono generati attraverso la funzione Generate_BlockDiagonal.

```{r message=FALSE, warning=FALSE}
# Import simulated data
BaseMat = as.matrix(read_csv('simulated_data/BaseMat.csv'))        # matrix Phi, dim = 200 x 40
y_hat_true = as.matrix(read_csv('simulated_data/y_hat_true.csv'))  # y true, dim = 300 x 200
beta_true = as.matrix(read_csv('simulated_data/beta_true.csv'))    # beta, dim = 300 x 40
mu_true = as.matrix(read_csv('simulated_data/mu_true.csv'))        # mu true, dim = 40 x 1

# object containing the true graph G, the true precision matrix K
simKG <- readRDS("simulated_data/simKG.rds")                       

n <- dim(y_hat_true)[1]  # n=300
r <- dim(y_hat_true)[2]  # r=200
p <- dim(BaseMat)[2]     # p=40
```

# Initialization

```{r}
# Compute quantities for function UpdateParamGSL
tbase_base = t(BaseMat)%*%BaseMat             # p x p (phi_t * phi)
tbase_data = t(BaseMat)%*%t(y_hat_true)       # p x n (phi_t * Y_t)  
Sdata = sum(diag(y_hat_true%*%t(y_hat_true))) # initialize phi*beta = 0 

set_UpdateParamsGSL_list = set_UpdateParamsGSL (
    tbase_base  = tbase_base,
    tbase_data  = tbase_data,
    Sdata       = Sdata,
    a_tau_eps   = 2000,
    b_tau_eps   = 2,
    sigma_mu    = 100,
    r           = r,
    Update_Beta = TRUE,
    Update_Mu   = TRUE,
    Update_Tau  = TRUE
)

# Set the number of iterations and burn-in
niter <- 10000
burn_in <- 1000

# Set the initialization values of the chains
initialization_values = set_initialization(
    Beta          = matrix(rnorm(n = p*n), nrow = p, ncol = n),
    mu            = rnorm(n=p),
    tau_eps       = 100,
    K             = simKG$Prec,
    G             = simKG$Graph,
    z             = c(rep(1,13), rep(2,13), rep(3,14)), 
    rho           = c(13, 13, 14),
    sigma         = 0.5,
    theta         = 1,
    weights_a0    = rep(1,p-1),
    weights_d0    = rep(1,p-1),
    total_weights = 0,
    total_K       = simKG$Prec,
    total_graphs  = simKG$Graph,
    graph_start   = NULL,
    graph_density = sum(simKG$Graph) / (p*(p-1)),
    beta_sig2     = 0.2,
    d             = 3
)

```

# Gibbs sampler

```{r}
chains = Gibb_sampler_update(
  set_UpdateParamsGSL_list,
  niter, 
  initialization_values,
  alpha_target           = 0.234,
  alpha_add              = 0.5,
  adaptation_step        = 1/(p*1000),
  seed                   = 123456,
  update_sigma_prior     = FALSE,
  update_theta_prior     = FALSE,
  update_weights         = FALSE,
  update_partition       = FALSE,
  update_graph           = FALSE,
  perform_shuffle        = FALSE
)
```

# Useful plots: 1. Plot smoothed curves

```{r}
# Compute the mean of Beta in order to have data_post
sum_Beta <- matrix(0, p, n)
for(i in (burn_in+1):niter){
  sum_Beta <- sum_Beta + chains$Beta[[i]]
}
mean_Beta <- sum_Beta/(niter-burn_in)
data_post <- BaseMat %*% mean_Beta

# Compute the x value, create the basis and the functional object
x <- seq(0, 1, length.out=r)
basis <- create.bspline.basis(rangeval=range(x), nbasis=40, norder=3) 
data.fd <- Data2fd(y = data_post, argvals = x, basisobj = basis)

# Plot smoothed curves
plot.fd(data.fd[1,], main="smoothed curves", ylim=c(-2,4))
# plot(x, data_post[,1], type='l', ylim=c(-2,4))
lines(x,y_hat_true[1,], main="smoothed curves", col='red')

```

# Useful plots: 2. Plot of the final Beta matrix

```{r}
ACutils::ACheatmap(
  chains$Beta[[niter]],  
  use_x11_device = F,
  horizontal = F,
  main = "Estimated Beta matrix",
  center_value = NULL,
  col.upper = "black",
  col.center = "grey50",
  col.lower = "white"
)

```

# Useful plots: 2. Traceplots (tau_eps, mu, beta)

```{r}
# tau_eps
tau_plot <- as.vector(chains$tau_eps)
tau_plot <- tau_plot[(burn_in+1):niter]
plot(ts(mcmc(tau_plot)), ylim=c(95,105))
abline(h=tau_plot[1], col='red', lwd=3)


# mu
mu_plot <- matrix(0, niter, p)
for(i in 1:niter){
  mu_plot[i, ] <- chains$mu[[i]]
}
par(mfrow=c(1,3))
plot(ts(mcmc(mu_plot[, 1])))
abline(h=mu_true[1], col='red')
plot(ts(mcmc(mu_plot[, 2])))
abline(h=mu_true[2], col='red')
plot(ts(mcmc(mu_plot[, 3])))
abline(h=mu_true[3], col='red')


# first element of first beta
beta1_plot <- rep(0, niter)
for(i in 1:niter){
  beta1_plot[i] <- chains$Beta[[i]][1,1]
}
# and 10th element of first beta
beta10_plot <- rep(0, niter)
for(i in 1:niter){
  beta10_plot[i] <- chains$Beta[[i]][10,1]
}

par(mfrow=c(1,2))
plot(ts(mcmc(beta1_plot)))
abline(h=beta_true[1,1], col='red')

plot(ts(mcmc(beta10_plot)))
abline(h=beta_true[1,10], col='red')


```
